\LARGE\singlespacing{\textbf{S1 Appendix} \\ \medskip Spatiotemporal modeling of \emph{Mycobacterium tuberculosis} transmission risks using environmental, clinical, and patient movement data}

### 2023-08-03 doc/paper/supplements.tex

\title{\LARGE\singlespacing{\textbf{S1 Appendix} \\ \medskip
Spatiotemporal modeling of \emph{Mtb} transmission in a South African primary care clinic}}

@@

The unknown parameter $q$ is not the actual number of infectious particles in the air. It rather represents the dose of infectious particles that corresponds to a certain probability of infection with a Poisson relationship, \eg one quantum corresponds to $P = 1 - \exp (1) = 63\%$ risk of infection\cite{Rudnick2003IndoorAir}. It is an attempt to consider the stochastic behavior of airborne transmission, since the number of infectious particles that cause an infection is unknown. Moreover, the risk of transmission depends on the environment, characteristics of the pathogen, and characteristics of the individual. For example, the type of respiratory (\eg breathing, coughing, or sneezing) of the infectious individual influences the size distribution of the generated particles, with smaller particles getting more easily deposited in the lower respiratory tract of the susceptible person\cite{Wang2021Science}, increasing the probability of infection.

@@

While the extension by Rudnick and Milton resolves the steady-state assumption, it still assumes a well-mixed airspace. Therefore, our aim is to extend the Wells-Riley model further to allow the risk of infection to vary both across time and airspace. We accomplish this by modeling the quanta concentration spatiotemporally. The risk of infection then is the cumulative exposure to infectious quanta over time and space, which depends on the susceptible individual's time-varying positions in the indoor space and quanta concentration at these time points. Formally, we estimate the cumulative risk of infection as 
\begin{equation}\label{eq:spattemp-P}
    P = \sum_s \sum_t N_{s,t} \cdot p,
\end{equation}
where $N$ is the quanta concentration in unit airspace $s$ at time $t$ (in quanta/$m^3$), and $p$ is the breathing rate per person (in $m^3/s$). The spatial quanta concentration changes over time according to the generation, diffusion and removal of quanta: 
\begin{itemize}
    \item[\ref{sec:quanta-generation}] Quanta generation: Infectious individuals generate quanta at their location in the indoor space. 
    \item[\ref{sec:quanta-diffusion}] Quanta diffusion: Over time, the infectious quanta diffuse and we approach a well-mixed airspace. 
    \item[\ref{sec:quanta-removal}] Quanta removal: Infectious quanta is removed from the air through (a)~outdoor air exchange, (b)~viral inactivation, and (c)~gravitational settling. 
\end{itemize}
We describe each process in more detail in the following. While we apply our model to \emph{Mtb}, we note that it can be applied to airborne respiratory infections in general.

@@

\subsection{Building and setup}

@@

An analysis of exhalation activities suggest a propagation distance (\ie the maximum distance from the source to the end of the exhaled air before dispersion) of 0.7m\cite{Tang2013PLoSOne}, but larger distances may be reached depending on the activity. Face masks reduce the propagation distance as most exhaled air (and particles) by locking the formation of the jet and leaking air on the sides of the masks\cite{Tang2009RoyalInt,Hui2012PLoSOne}. Therefore, we assume that mask wearing confined the quanta generation only to the cell where the infectious individual is located. This corresponds to a propagation distance of 0.35m (the diagonal cell length) and an area of 0.25$^2$m$^2$.

@@

\begin{figure}[!htpb]
    \centering
    \includegraphics{illustrations/prior-q.pdf}
    \caption{Prior distribution function for the quanta generation rate (quanta/h) along with reported estimates in the literature: 0.89 by Andrews (2014) [REF], 1.25 by Riley (1962) [REF], 8.2 by Escombe (2008) [REF], and 13 by Nardell (1991) [REF].}
    \label{fig:quanta-distribution}
\end{figure}

@@

Face masks reduce the number of viral-laden aerosols that are exhaled into the air\cite{Milton2013PLoSPathogens} and thus lower the generation of infectious quanta. The extent of the reduction is subject to uncertainty. A study with multidrug-resistant TB cases estimated a decrease of 56\% (95\%-CrI, 33\% to 70.5\%) when patients used surgical face masks\cite{Dharmadhikari2012AJRCCM}. This estimate was used by a modelling study to estimate the impact of mask wearing on \emph{Mtb} transmission in a primary care clinic. Therefore,, we model the reduction in infectious quanta through mask wearing with a Normal($\mu$ = 0.56, $\sigma$ = 0.11) truncated at 0 and 1. The effective quanta generation rate is thus $q$ times the reduction due to mask wearing. Again, we model what the risk of infection would have been without mask wearing by removing the reduction term from the model.

@@

\begin{figure}[!htpb]
    \centering
    \includegraphics{illustrations/prior-lambda.pdf}
    \caption{Prior distribution function for the viral inactivation rate along with reported estimates in the literature: 0.12 by Loudon (1969) [REF], 0.46 by Gannon (2007) [REF], 3.2 by Klein (2014) [REF], and 8.3 by Lever (2000) [REF].}
    \label{fig:lambda-distribution}
\end{figure}

>>>>

\title{\LARGE\singlespacing{\textbf{S1 Appendix} \\ \medskip
Spatiotemporal modeling of \emph{Mycobacterium tuberculosis} transmission risks using environmental, clinical, and patient movement data}}

@@

The unknown parameter $q$ is not the actual number of infectious particles in the air. It rather represents the dose of infectious particles that corresponds to a certain probability of infection with a Poisson relationship, \eg one quantum corresponds to $P = 1 - \exp (1) = 63\%$ risk of infection\cite{Rudnick2003IndoorAir}. It is an attempt to consider the stochastic behavior of airborne transmission, since the number of infectious particles that cause an infection is unknown. Moreover, the risk of transmission depends on the environment, characteristics of the pathogen, and characteristics of the individual. For example, the type of respiratory activity (\eg breathing, coughing, or sneezing) determines the site in the respiratory tract where the exhaled particles are generated, which in turn influences the particle size distribution\cite{Wei2016AMJIC}. Smaller particles reach deeper into the lungs of the susceptible person\cite{Wang2021Science}, which increases the probability of infection.

@@

While the extension by Rudnick and Milton resolves the steady-state assumption, it still assumes a well-mixed airspace. Therefore, our aim is to extend the Wells-Riley model further to allow the risk of infection to vary both across time and airspace. We accomplish this by modeling the quanta concentration spatiotemporally. The risk of infection then is the cumulative exposure to infectious quanta over time and space, which depends on the susceptible individual's time-varying positions in the indoor space and quanta concentration at these time points. Formally, we estimate the cumulative risk of infection for susceptible individual $a$ as 
\begin{equation}\label{eq:spattemp-P}
    P_a = \sum_s \sum_t N_{s,t} \cdot \mathbb{I}_{s,t}^a \cdot p,
\end{equation}
where $N$ is the quanta concentration in unit airspace $s$ at time $t$ (in quanta/$m^3$), $\mathbb{I}$ is a binary variable indicating whether $a$ was at location $s$ at time $t$, and $p$ is the breathing rate per person (in $m^3/s$). The spatial quanta concentration changes over time according to the generation, diffusion and removal of quanta: 
\begin{itemize}
    \item[\ref{sec:quanta-generation}] Quanta generation: Infectious individuals generate quanta at their location in the indoor space. 
    \item[\ref{sec:quanta-diffusion}] Quanta diffusion: Over time, the infectious quanta diffuse and we approach a well-mixed airspace. 
    \item[\ref{sec:quanta-removal}] Quanta removal: Infectious quanta is removed from the air through (a)~outdoor air exchange, (b)~viral inactivation, and (c)~gravitational settling. 
\end{itemize}
We describe each process in more detail in the following. While we apply our model to \emph{Mtb}, we note that it can be applied to airborne respiratory infections in general.

@@

\subsection{Building and setup}\label{prep:building}

@@

An analysis of exhalation activities suggest a propagation distance (\ie the maximum distance from the source to the end of the exhaled air before dispersion) of 0.7m\cite{Tang2013PLoSOne}, but larger distances may be reached depending on the activity. Face masks reduce the propagation distance both by blocking (the mask capturing particles) and deflecting particles (leakage of particles on the sides of the mask)\cite{Tang2009RoyalInt,Hui2012PLoSOne,Mansour2013AerosolMed}. Therefore, we assume that mask wearing confined the quanta generation only to the cell where the infectious individual is located. This corresponds to a propagation distance of 0.35m (the diagonal cell length) and an area of 0.25$^2$m$^2$.

@@

\begin{figure}[!htpb]
    \centering
    \includegraphics{illustrations/prior-q.pdf}
    \caption{Prior distribution function for the quanta generation rate (quanta/h) along with reported estimates in the literature: 0.89 by Andrews (2014)\cite{Andrews2014JID}, 1.25 by Riley (1962)\cite{Riley1962ARRD}, 8.2 by Escombe (2008)\cite{Escombe2008PLoSMed}, and 13 by Nardell (1991)\cite{Nardell1991ARRD}.}
    \label{fig:quanta-distribution}
\end{figure}

@@

Face masks reduce the number of viral-laden aerosols that are exhaled into the air\cite{Milton2013PLoSPathogens,Leung2020NatMed} and should thus lower the generation of infectious quanta. The extent of the reduction is subject to uncertainty. A study with multidrug-resistant TB cases estimated a decrease of 56\% (95\%-CrI, 33\% to 70.5\%) when patients used surgical face masks\cite{Dharmadhikari2012AJRCCM}. This estimate was used by a modelling study to estimate the impact of mask wearing on \emph{Mtb} transmission in a primary care clinic. Therefore,, we model the reduction in infectious quanta through mask wearing with a Normal($\mu$ = 0.56, $\sigma$ = 0.11) truncated at 0 and 1. The effective quanta generation rate is thus $q$ times the reduction due to mask wearing. Again, we model what the risk of infection would have been without mask wearing by removing the reduction term from the model.

@@

\begin{figure}[!htpb]
    \centering
    \includegraphics{illustrations/prior-lambda.pdf}
    \caption{Prior distribution function for the viral inactivation rate for \emph{Mtb tuberculosis} along with reported estimates in the literature: 0.12 by Loudon (1969)\cite{Loudon1969AMRRD}, 0.46 by Gannon (2007) for \emph{Mtb bovis}\cite{Gannon2007ResVetSci}, 3.2 by Klein (2014)\cite{Klein2014IJMyco}, and 8.3 by Lever (2000)\cite{Lever2000LettersAppliedMicrobio}.}
    \label{fig:lambda-distribution}
\end{figure}
